D3064R1: How to Avoid OOTA Without Really Trying

### 2024-05-12 ootavacuous.tex
Invented loads: Justify restrictions based on abstract machine
Suggested-by: Alan Stern <stern@rowland.harvard.edu>
Signed-off-by: Paul E. McKenney <paulmck@kernel.org>

However, an especially devious compiler might transform the source into
the following form before translating it to machine code:
\begin{quote}
\begin{verbatim}
 1  int r1;
 2  int r1a = (x != 0);
 3  int r1b = (x != 0);
 4  int r2 = (y != 0);
 5  if (r1a != r1b) {
 6      r1 = r2;
 7      z = 1;
 8  } else {
 9      r1 = r1b;
10      z = (r1 == r2);
11  }
\end{verbatim}
\end{quote}
The idea is that \co{r1a}, \co{r1b}, and \co{r2} can each be only zero or one,
so if \co{r1a} and \co{r1b} differ then one of them must be equal to \co{r2}.
In executions where this happens---because another thread writes to \co{x}
between the two loads---the implementation can choose at runtime to use
for \co{r1} whichever value agrees with \co{r2}, as shown on line~6.
Then the value stored to \co{z} on line~7 will always be one,
with no dependence on the value loaded from \co{y}.
This example is discussed more fully in
Appendix~\ref{app:Inventing Atomic Loads}.

@@

\paragraph{Invented Atomic Loads}
\begin{listing}[tbp]
\scriptsize
\begin{verbatim}
 1 struct foo {
 2   int a;
 3   int b;
 4 };
 5 _Atomic struct foo *globalfoop;
 6
 7 void bar()
 8 {
 9   int a;
10   int b;
11   struct foo *fp;
12
13   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
14   a = fp->a;
15   // fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
16   b = fp->b;
17   do_something_with(a, b);
18 }
\end{verbatim}
\caption{Example Duplicated Load}
\label{lst:Example Duplicated Load}
\end{listing}
Although one can argue that it is functionally correct to invent atomic
loads as long as the loaded value is not used, doing so can result in
extra cache misses for no good purpose.
And duplicating an atomic load can also lead to incorrect results if done
carelessly.
To see this, consider
Listing~\ref{lst:Example Duplicated Load}.
Suppose that the load on line~13 were duplicated, for example, by
uncommenting line~15.
Then the values of \co{a} and \co{b} passed to line~17's call to
\co{do_something_with()} might be from different instances of the
\co{foo} structure, something that \co{do_something_with()} probably
isn't prepared to deal with.

@@

There are situations where inventing loads from non-atomic objects
is customary, for example, when hoisting a load out of a loop.
In this case, compilers might hoist the load from \co{x} out of
the loop:
\begin{quote}
\begin{verbatim}
do_something_with(x);
for (i = 0; i < limit; i++)
    y[i] += x;
\end{verbatim}
\end{quote}
A compiler might act as if the program had instead been this:
\begin{quote}
\begin{verbatim}
r1 = x;
do_something_with(r1);
for (i = 0; i < limit; i++)
    y[i] += r1;
\end{verbatim}
\end{quote}
Note that passing \co{x} to \co{do_something_with()} is important
when \co{x} is not atomic,
because this allows the compiler to assume that inventing additional
accesses to \co{x} will not result in a data race.
Without this assumption, the above transformation might introduce a data
race when \co{limit} is less than or equal to zero.
Hence the use of quotation marks in the section heading:
Here the load is not actually invented, but instead merged with an
earlier load with no synchronization between the two loads.

@@

In addition, compiler writers should carefully consider the issues
raised by the following section:

@@

Consider the code below:
\begin{quote}
\begin{verbatim}
int r0 = x;
int r1 = r0 * r0 + 2 * r0 + 1;
\end{verbatim}
\end{quote}
A developer might be forgiven for assuming that the value of \co{r1}
is guaranteed to be a perfect square, at least for sufficiently small
values of \co{x}.

But if the compiler is permitted to invent atomic loads, there is no such
guarantee, as can be seen from this transformed version of the above code:
\begin{quote}
\begin{verbatim}
int r0 = x;
int invented = x;
int r1 = r0 * r0 + 2 * invented + 1;
\end{verbatim}
\end{quote}
For example, if the load into \co{r0} happened when the value of \co{x}
was zero and the load into \co{invented} happened when the value of
\co{x} was ten, then the value of \co{r1} will be twenty, which is not
a perfect square.

Similar issues can arise even when the compiler merely repurposes a
value from an atomic load, as can be seen from this example:
\begin{quote}
\begin{verbatim}
do_something(x);
int r0 = x;
int r1 = r0 * r0 + 2 * r0 + 1;
\end{verbatim}
\end{quote}
Again, a developer might be forgiven for assuming that the value of
\co{r1} is guaranteed to be a perfect square, at least for sufficiently
small values of \co{x}.
But is the compiler allowed to transform this example into the following?
\begin{quote}
\begin{verbatim}
repurposed = x;
do_something(repurposed);
int r0 = x;
int r1 = r0 * r0 + 2 * repurposed + 1;
\end{verbatim}
\end{quote}
Again, this can result in the value of \co{r1} being something other than
a perfect square.
%
But it would be just fine if the compiler fully merged the two loads from
\co{x}, for example, as follows:
\begin{quote}
\begin{verbatim}
int r0 = x;
do_something(r0);
int r1 = r0 * r0 + 2 * r0 + 1;
\end{verbatim}
\end{quote}

Similar observations apply to atomic loads from pointers.
For example, consider
Listing~\ref{lst:Example Duplicated Load} on
page~\pageref{lst:Example Duplicated Load},
in which the developer expected line~19 to pass two fields from the
same structure to \co{do_something_with()}, which is an entirely
reasonable expectation.
This expectation will not always be met if the compiler is permitted
to invent atomic loads as shown by the commented-out line~15.
\begin{listing}[tbp]
\scriptsize
\begin{verbatim}
 1 struct foo {
 2   int a;
 3   int b;
 4 };
 5 _Atomic struct foo *globalfoop;
 6
 7 void bar()
 8 {
 9   int a;
10   int b;
11   struct foo *fp;
12
13   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
14   do_something(fp);
15   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
16   a = fp->a;
17   // fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
18   b = fp->b;
19   do_something_with(a, b);
20 }
\end{verbatim}
\caption{Example Merging of Duplicated Loads}
\label{lst:Example Merging of Duplicated Loads}
\end{listing}
But suppose that there was an earlier load from \co{globalfoop}, as
shown on lines~13 and~14 in
Listing~\ref{lst:Example Merging of Duplicated Loads}.
Then, assuming that \co{do_something()} is synchronization-free,
the second load on line~15 could be removed without violating user
expectations.

It is imperative that users be able to rely on arithmetic and pointer accesses.
Therefore, compilers must refrain from inventing atomic loads and from repurposing
their values, but compilers are permitted to merge adjacent atomic loads from the
same variable.
These restrictions correspond to our notion of quasi volatile defined in
Section~\ref{sec:Volatile and Quasi Volatile Accesses}.

>>>>

However, an especially devious compiler might transform the source into
the following form before translating it to machine code:
\begin{quote}
\begin{verbatim}
 1  int r1;
 2  int r1a = (x != 0);
 3  int r1b = (x != 0);
 4  int r2 = (y != 0);
 5  if (r1a != r1b) {
 6      r1 = r2;
 7      z = 1;
 8  } else {
 9      r1 = r1b;
10      z = (r1 == r2);
11  }
\end{verbatim}
\end{quote}
The idea is that \co{r1a}, \co{r1b}, and \co{r2} can each be only zero or one,
so if \co{r1a} and \co{r1b} differ then one of them must be equal to \co{r2}.
In executions where this happens---because another thread writes to \co{x}
between the two loads---the implementation can choose at runtime to use
for \co{r1} whichever value agrees with \co{r2}, as shown on line~6.
Then the value stored to \co{z} on line~7 will always be one,
with no dependence on the value loaded from \co{y}.
This example is discussed more fully in
Appendix~\ref{app:Inventing Atomic Loads},
which also discusses other examples in which invented atomic loads lead
to incorrect behavior, that is, behavior that is not exhibited by the
abstract machine.\footnote{
	See also the ``Invented Atomic Loads'' paragraph of
	Section~\ref{sec:Constraints of the Standard}.}

@@

\paragraph{Invented Atomic Loads}
\begin{listing}[tbp]
\scriptsize
\begin{verbatim}
 1 struct foo {
 2   int a;
 3   int b;
 4 };
 5 _Atomic struct foo *globalfoop;
 6
 7 void bar()
 8 {
 9   int a;
10   int b;
11   struct foo *fp;
12
13   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
14   a = fp->a;
15   // fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
16   b = fp->b;
17   do_something_with(a, b);
18 }
\end{verbatim}
\caption{Example Duplicated Load}
\label{lst:Example Duplicated Load}
\end{listing}
Although one can argue that it is functionally correct to invent atomic
loads as long as the loaded value is not used, doing so can result in
extra cache misses for no good purpose.
And duplicating an atomic load can also lead to incorrect results if done
carelessly.
To see this, consider
Listing~\ref{lst:Example Duplicated Load}.
Suppose that the load on line~13 were duplicated, for example, by
uncommenting line~15.
Then the values of \co{a} and \co{b} passed to line~17's call to
\co{do_something_with()} might be from different instances of the
\co{foo} structure, something that \co{do_something_with()} probably
isn't prepared to deal with and that represents incorrect behavior,
that is, behavior that the abstract machine does not exhibit.
See
Appendix~\ref{app:Inventing Atomic Loads}
for more details on this example and other examples of invented
atomic loads.

@@

There are situations where inventing loads from non-atomic objects
is customary, for example, when hoisting a load out of a loop.
In this case, compilers might hoist the load from \co{x} out of
the loop:
\begin{quote}
\begin{verbatim}
do_something_with(x);
for (i = 0; i < limit; i++)
    y[i] += x;
\end{verbatim}
\end{quote}
A compiler might act as if the program had instead been this:
\begin{quote}
\begin{verbatim}
r1 = x;
do_something_with(r1);
for (i = 0; i < limit; i++)
    y[i] += r1;
\end{verbatim}
\end{quote}
Note that passing \co{x} to \co{do_something_with()} is important
when \co{x} is not atomic,
because this allows the compiler to assume that inventing additional
accesses to \co{x} will not result in a data race.
Without this assumption, the above transformation might introduce a data
race when \co{limit} is less than or equal to zero.
Hence the use of quotation marks in the section heading:
Here the load is not actually invented, but instead merged with an
earlier load such that there is no synchronization between the two loads.

@@

In addition, compiler writers should carefully consider the issues
raised by the following section.

@@

Consider the code below:
\begin{quote}
\begin{verbatim}
int r0 = x;
int r1 = r0 * r0 + 2 * r0 + 1;
\end{verbatim}
\end{quote}
For sufficiently small values of \co{x}, the abstract machine is
guaranteed to produce a perfect square in \co{r1}.

But if the compiler is permitted to invent atomic loads, then this
guarantee might not hold, as can be seen from this transformed version of
the above code:
\begin{quote}
\begin{verbatim}
int r0 = x;
int invented = x;
int r1 = r0 * r0 + 2 * invented + 1;
\end{verbatim}
\end{quote}
For example, if the load into \co{r0} happened when the value of \co{x}
was zero and the load into \co{invented} happened when the value of
\co{x} was 10, then the value of \co{r1} will be 21, which is not
a perfect square.

Similar issues can arise even when the compiler merely repurposes a
value from an atomic load, as can be seen from this example:
\begin{quote}
\begin{verbatim}
do_something(x);
int r0 = x;
int r1 = r0 * r0 + 2 * r0 + 1;
\end{verbatim}
\end{quote}
Again, for sufficiently small values of \co{x}, the abstract machine is
guaranteed to produce a perfect square in \co{r1}.
So is the compiler allowed to transform this example into the following?
\begin{quote}
\begin{verbatim}
repurposed = x;
do_something(repurposed);
int r0 = x;
int r1 = r0 * r0 + 2 * repurposed + 1;
\end{verbatim}
\end{quote}
This can result in the value of \co{r1} being something other than
a perfect square, which is incorrect behavior.
%
But it would be just fine if the compiler fully merged the two loads from
\co{x}, for example, as follows:
\begin{quote}
\begin{verbatim}
int r0 = x;
do_something(r0);
int r1 = r0 * r0 + 2 * r0 + 1;
\end{verbatim}
\end{quote}

Similar observations apply to atomic loads from pointers.
For example, consider
Listing~\ref{lst:Example Duplicated Load} on
page~\pageref{lst:Example Duplicated Load},
in which the abstract machine guarantees that line~17 passes two
fields from the same structure to \co{do_something_with()}.
This guarantee will not always be met if the compiler is permitted to
invent atomic loads such as that contained in the commented-out line~15.
\begin{listing}[tbp]
\scriptsize
\begin{verbatim}
 1 struct foo {
 2   int a;
 3   int b;
 4 };
 5 _Atomic struct foo *globalfoop;
 6
 7 void bar()
 8 {
 9   int a;
10   int b;
11   struct foo *fp;
12
13   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
14   do_something(fp);
15   fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
16   a = fp->a;
17   // fp = atomic_load_explicit(&globalfoop, memory_order_acquire);
18   b = fp->b;
19   do_something_with(a, b);
20 }
\end{verbatim}
\caption{Example Merging of Duplicated Loads}
\label{lst:Example Merging of Duplicated Loads}
\end{listing}
But suppose that there was an earlier load from \co{globalfoop}, as
shown on lines~13 and~14 in
Listing~\ref{lst:Example Merging of Duplicated Loads}.
Then, assuming that \co{do_something()} is synchronization-free, the
second load on line~15 could be removed without resulting in incorrect
behavior.

Please note that these examples are not just accidents of the current state
of the C++ abstract machine.
After all, it is imperative that users be able to rely on the laws of
arithmetic and on atomic accesses to pointers.
Therefore, compilers must refrain from inventing atomic loads and from repurposing
their values.
However, compilers are permitted to merge adjacent atomic loads from
the same variable.
These restrictions correspond to our notion of quasi volatile defined in
Section~\ref{sec:Volatile and Quasi Volatile Accesses}.
